\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-cancer chemotherapeutic sensitivity \\

### 2020-06-01 manuscript/manuscript.tex
first rough draft done

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic response matrices obtained from the GDSC consortium were used to evaluate conventional tools used to separate patients into chemotherapeutic response groups as well as bifurcate candidate cell lines into defined response cohorts. Classical cancer classifications, including anatomic location and solid vs. non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status were used to evaluate whether classical classification systems were able to stratify cell lines into chemotherapeutic response groups (\cite{colorectal}, \cite{gi}, \cite{lung}, \cite{breast}). Cell-lines were separated into subgroups based on anatomic location, solid vs. non-solid tumour status, and mutation/non-mutation groups for TP53 and KRAS respectively. These subgroups were then plotted with the first and second principal components to identify if they clustered together. Observed clustering would be indicative of these classifiers being predictive of chemotherapeutic response.

@@

% Discussion
\section{Discussion}
In this study we developed an accurate deep learning-based model for classifying cancer patients based on RNA expression data into therapeutic response cohorts on a relatively small cell line therapeutic response dataset. Using unsupervised clustering techniques, we segregated cancer patients into two defined therapeutic response groups with significantly different responses to a multitude of standard chemotherapies. [EXPAND HERE]

The relatively low accuracy of the neural networks with 1 and 2 hidden layers (82.6\% and 70.8\% respectively) suggests that the therapeutic response cohorts cannot be separated by a linear classifier. Furthermore, the high FNR of the network with 5 hidden layers as compared to the 3 and 4 hidden layer networks indicates overfitting. To this end, either a 3 or 4 hidden layer network is the ideal architecture for analyzing our data. Interestingly, there is no significant difference between AUC for these models (Fig. \ref{fig:roc}). This indicates that the major factor for model evaluation above two hidden layers for this dataset are the FPR and FNR of the classifications.

A major limitation of our study was the availability of large datasets to train our model. Here we faced a $p$ $\gg$ $n$ problem as machine learning models expect that the number of features $p$ is much larger than the number of observations $n$. To minimize this bias, we applied the Boruta algorithm to reduce our 16,382 genes by 541 cell lines dataset to a 301 by 541 matrix. The algorithm has been shown in various journals to be an effective feature selector method in high dimensional omics datasets \cite{boruta}. To prevent overfitting of the reduced matrix, we applied batch normalization and dropout layers immediately preceding each hidden layer. Of note, a neural network architecture where the initial hidden layers diverge and the latter hidden layers converge provides the most accurate classifications of the RNA-seq data.

Future investigations will look to validate the efficacy of the model in prediction of chemotherapy response in various forms of cancer using the current transcriptomic signature. It is likely that the model accuracy will vary between therapy targets, as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. Furthermore, it may be interesting to select relevant genes using a different feature selector. The selected algorithm in this paper, Boruta, operates on patterns of statistical relationships rather than biological relationships. Our use of Boruta was inspired by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors \cite{boruta, deep_cell}. Computational efficiency and the resulting feature set quality were also motivators for choosing BorutaPy over other selection algorithms such as univariate selection or principle component analysis. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.

@@

\begin{table*}[!ht]
	\caption{Neural network architectures and the number of neurons per layer ($n=560$).}	\label{tab:params}
	\centering
	\begin{tabular}{l l}
		\toprule
		Architecture & Number of neurons \\
		\midrule
		 5 hidden layers & 300 inputs, 2n, 4n, 4n, 2n, n, binary output \\
		 4  hidden layers & 300 inputs, 2n, 4n, 4n, 2n, binary output \\
		 3 hidden layers & 300 inputs, 2n, 4n, 2n, binary output \\
		\bottomrule
	\end{tabular}
\end{table*}

>>>>

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic response matrices obtained from the GDSC consortium were used to evaluate conventional tools used to separate patients into chemotherapeutic response groups as well as bifurcate candidate cell lines into defined response cohorts. Classical cancer classifications, including anatomic location and solid vs. non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status were used to evaluate whether classical classification systems were able to stratify cell lines into chemotherapeutic response groups \cite{colorectal}, \cite{gi}, \cite{lung}, \cite{breast}. Cell-lines were separated into subgroups based on anatomic location, solid vs. non-solid tumour status, and mutation/non-mutation groups for TP53 and KRAS respectively. These subgroups were then plotted with the first and second principal components to identify if they clustered together. Observed clustering would be indicative of these classifiers being predictive of chemotherapeutic response.

@@

% Discussion
\section{Discussion}
It is well known that there is substantial heterogeneity with respect to chemotherapeutic response between, and among, cancer types. Although there have been multiple attempts to identify molecular and clinical features predictive of response to particular targeted therapies, there remains considerable variability within subgroups identified using these factors. In this study, we attempt to accurately stratify cancer cell lines into defined groups based on response to a large range of chemotherapeutics, and to create a deep learning transcriptomic model capable of accurately categorizing samples into these defined groups. Using cell-line chemotherapeutic efficacy data obtained from the GDSC, we employ unsupervised clustering techniques to identify two defined therapeutic response groups with significantly different responses to a multitude to standard chemotherapies. We show that these clusters outperform classical methods of stratifying samples into chemotherapeutic categories based on clinical and molecular criteria. Here, we use a biologically agnostic feature selection algorithm, Boruta, to reduce the original set of 16,382 genes to a subset of 300 genes and their associated IC$_{50}$. These were fed into neural networks, which were then optimized. We determined from confusion matrices and ROC curve analysis that the best network architecture included 4 hidden layers.

The comparatively lower accuracy of the neural networks with 1 and 2 hidden layers (82.6\% and 70.8\% respectively) suggests that the therapeutic response cohorts cannot be separated by a linear classifier. Furthermore, the high FNR of the network with 5 hidden layers as compared to the 3 and 4 hidden layer networks indicates overfitting. To this end, either a 3 or 4 hidden layer network is the ideal architecture for analyzing our data. Interestingly, there is no significant difference between AUC for these models (Fig. \ref{fig:roc}). This indicates that the major factor for model evaluation above two hidden layers for this dataset are the FPR and FNR of the classifications.

The deep learning transcriptomic model consists of 300 genes, with KEGG pathway enrichment suggesting that predictive genes are associated with numerous pathways, most notably PI3K signalling and focal adhesion pathways. Interestingly, there is a growing body of literature that suggests that PI3K pathway dysregulation may be associated with chemotherapeutic resistance in numerous different cancer and treatment contexts \cite{huang_2009}. Numerous studies have identified increases in Akt signalling in cancer cell lines exposed to chemotherapy and radiotherapy. Moreover, significant increases in Akt have been identified in chemoresistant and radioresistant cancer models. Similarly, several studies have identified focal adhesion as a potential protective mechanism for various cancer cells \cite{focal_adhesion}. In fact, inhibition of particular integrin isoforms has been shown to increase the susceptibility of various cancer cell lines to conventional chemo/radiotherapies. Our results provide further evidence that dysregulation of PI3K signalling and focal adhesion may play a role in chemotherapy resistance in a pan-cancer context.

A major limitation of our study was the availability of large datasets to train our model. Here we faced a $p$ $\gg$ $n$ problem as machine learning models expect that the number of features $p$ is much larger than the number of observations $n$. To minimize this bias, we applied the Boruta algorithm to reduce our 16,382 genes by 541 cell lines dataset to a 300 by 541 matrix. The algorithm has been shown in various journals to be an effective feature selector method in high dimensional omics datasets (\cite{boruta}). To prevent overfitting of the reduced matrix, we applied batch normalization and dropout layers immediately preceding each hidden layer. Of note, a neural network architecture where the initial hidden layers diverge and the latter hidden layers converge provides the most accurate classifications of the RNA-seq data.

Future investigations will look to validate the efficacy of the model in prediction of chemotherapy response in various forms of cancer using the current transcriptomic signature. It is likely that the model accuracy will vary between therapy targets, as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. Furthermore, it may be interesting to select relevant genes using a different feature selector. Our selected algorithm, Boruta, operates on patterns of statistical relationships rather than biological relationships. Our use of Boruta was inspired by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors (\cite{boruta}, \cite{deep_cell}). Computational efficiency and the resulting feature set quality were also motivators for choosing BorutaPy over other selection algorithms such as univariate selection or principle component analysis. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.

@@

\begin{table*}[!ht]
	\caption{Neural network architectures and the number of neurons per layer ($n=560$).}	\label{tab:architectures}
	\centering
	\begin{tabular}{l l}
		\toprule
		Architecture & Number of neurons \\
		\midrule
		 5 hidden layers & 300 inputs, 2n, 4n, 4n, 2n, n, binary output \\
		 4  hidden layers & 300 inputs, 2n, 4n, 4n, 2n, binary output \\
		 3 hidden layers & 300 inputs, 2n, 4n, 2n, binary output \\
		\bottomrule
	\end{tabular}
\end{table*}
